---
title: "Exploration of Fecal Indicator Bacteria data sources in the Tampa Bay Estuary Program watersheds"
subtitle: "Prepared for the Tampa Bay Estuary Program"
date: 2024-02-14
date-modified: today
author: 
    name: Kim Cressman
    orcid: 0000-0002-8083-2434
    email: kim@catbirdstats.com 
    affiliation: 
        name: Catbird Stats, LLC
        url: https://www.catbirdstats.com
format: 
    html:
        fig-width: 7
        fig-height: 7
        toc: true
embed-resources: false
execute:
    echo: false
    message: false
    warning: false
    error: true
params:
    nYearSegments: 5
    YeartoSplit: 2018
---

```{r}
library(tidyverse)
library(sf)
library(mapview)
library(khroma) # for colorblind-friendly color palettes
library(tbeptools)
library(kableExtra)
library(plotly)

options(knitr.duplicate.label = "allow")  # so I can still use chunk names in the child doc - useful for navigation
```

```{r read-shapefiles}
path <- here::here("GIS")

bay_segs <- st_read(here::here(path, "TBEP-Bay_Segments.shp"),
                    quiet = TRUE)
wbids <- st_read(here::here(path, "Waterbody_IDs_(WBIDs).shp"),
                    quiet = TRUE)

# tbseg is a polygon layer included in tbeptools
# tidalcreeks is as well but needs to be subsetted; covers most of SWFL
```

```{r modify-shapefiles}
# pull out crs of the wbids for future joins, and make sure bay_segs is in that crs
to_crs <- st_crs(wbids)
bay_segs <- st_transform(bay_segs, crs = to_crs)

# subset wbids to only those that intersect bay segments
wbids <- st_intersection(wbids, bay_segs)
```

```{r}
yrsplit <- params$YeartoSplit
tslengthsplit <- params$nYearSegments

col_scale_divs <- tslengthsplit  # will actually be one category fewer than this, because of the sequence endpoints
```

This document will compare a hard download through the Water Quality Portal's website to the results from `tbeptools::read_importfib()`. The same figures and tables are produced for both download sources. They are *not* presented as separate tabs because that removed the functionality of a table of contents.   


# 1.  Website download  


### FIB Sampling points

This file contains *results* from microbiological sampling in the 5 TBEP counties from 1995-2022. We don't necessarily need to know what the results are, just that they exist. The query from Water Quality Portal was also set to only include stations with at least 10 sampling events. The latest download was performed 1/29/2024 and [this is the link to the query](https://www.waterqualitydata.us/#countrycode=US&statecode=US%3A12&countycode=US%3A12%3A057&countycode=US%3A12%3A081&countycode=US%3A12%3A101&countycode=US%3A12%3A103&countycode=US%3A12%3A105&sampleMedia=Water&characteristicType=Microbiological&startDateLo=01-01-1995&startDateHi=12-31-2022&minactivities=10&mimeType=csv&dataProfile=biological&providers=NWIS&providers=STEWARDS&providers=STORET).

That download was pre-processed using `R/WQP_preprocessing.R` to:  

-  filter out Groundwater samples - only keeping Surface Water.  
-  only keep sample types Fecal Coliform, Total Coliform, Enterococcus, and E. coli (eliminates DNA-based samples). Total Coliform is in the processed data frame but I am ignoring it in this exploration.  
-  only keep a small subset of columns relevant to this exploration.  

The pre-processing reduced the file size from 167MB to 40MB, and the smaller file is used in this document.  

```{r read-sample-info}
fib_samps <- readRDS(here::here("data-WQP", "biological_reduced.rds")) |> 
    filter(CharacteristicName %in% c("Fecal Coliform", "Enterococcus", "Escherichia coli"))
```

```{r, child = "_fib-expl-sourced.qmd"}

```

```{r}
temporal_from_website <- entero_toMap
```


***
***

# 2.  `read_importwqp` download  


```{r read-sample-info2}
fib_samps <- readRDS(here::here("data-WQP", "tbeptools_importwqp.rds"))

# doing this to match names from web download
# probably would have handled differently if I'd downloaded via tbeptools earlier
names(fib_samps) <- c("OrgID", "SampleStation", "SampleDate",
                      "class", "yr", "mo", "Lat", "Long",
                      "Sample_Depth_m", "CharacteristicName",
                      "val", "uni", "qual")

org_matching <- tribble(
    ~"OrgID", ~"OrgName",
    "21FLDOH_WQX",   "Department of Health",
    "21FLHILL_WQX",  "Hillsborough County",
    "21FLMANA_WQX",  "Manatee County",
    "21FLPASC_WQX",  "Pasco County",
    "21FLPDEM_WQX",  "Pinellas County",
    "21FLPOLK_WQX",  "Polk County"
)

fib_samps <- fib_samps |> 
    filter(CharacteristicName %in% c("ecocci", "ecoli", "fcolif")) |> 
    mutate(SampleDate = as.Date(SampleDate),
           CharacteristicName = case_match(CharacteristicName,
                                           "ecocci" ~ "Enterococcus",
                                           "ecoli" ~ "Escherichia coli",
                                           "fcolif" ~ "Fecal Coliform",
                                           "totcol" ~ "Total Coliform",
                                           .default = "problem here!")) |> 
    left_join(org_matching)
```

```{r, child = "_fib-expl-sourced.qmd"}

```

```{r}
temporal_from_tbeptools <- entero_toMap
```


#  3. Comparison of downloads  

### Map of Enterococcus sampling stations by download method  

As with most maps above, only looking at waters in 2 and 3M classes.  

```{r}
mapview(bay_segs,
        col.regions = map_colors[4],
        layer.name = "Bay Segment") +
    mapview(temporal_from_tbeptools,
            layer.name = "stations in read_importwqp download") +
    mapview(temporal_from_website,
            layer.name = "stations in website download",
            col.regions = map_colors[5])
```


Most of the stations appeared in both downloads, because the bulk of the stations are sampled by organizations that were already written into `tbeptools::read_importwqp()`.  

**Stations in `read_importwqp` download but not website download:**  

-  Stations sampled by Pinellas County. The website download was limited to stations that were sampled at least 10 times, so the stratified random sampling setup of Pinellas County did not yield stations. These stations are certainly worth considering in a report card, as they go back to 2009.  


**Stations in website download but not `read_importwqp` download:**  

-  10 stations sampled by the City of St. Petersburg, OrgID 21FLCOSP_WQX - primarily in the Boca Ciega Bay segment. Data in the WQP spanned the range or 2016 or 2017 until 2021. 
-  13 stations sampled by the FL Department of Environmental Protection, Southwest District, OrgID 21FLTPA_WQX. There are 1-5 of these in each major bay segment. Sampling ended for some by 2017, but others were current through 2022. 


#  4. Recommendations  

Based on this exploration of FIB data availability, we make the following recommendations:  

1.  Use Enterococcus as the fecal indicator of choice in estuarine report cards. The temporal coverage for this indicator is almost universally more recent than for fecal coliform.  
2.  Add 21FLCOSP_WQX and 21FLTPA_WQX as valid entries for `org` in `tbeptools::read_importwqp()`.  



