---
title: "FIB Exploration"
format: 
    html:
        fig-width: 7
        fig-height: 7
        toc: true
embed-resources: false
execute:
    echo: false
    message: false
    warning: false
    error: true
params:
    nYearSegments: 5
    YeartoSplit: 2018
---

```{r}
library(tidyverse)
library(sf)
library(mapview)
library(khroma) # for colorblind-friendly color palettes
library(tbeptools)
library(kableExtra)
library(plotly)
```

```{r read-shapefiles}
path <- here::here("GIS")

bay_segs <- st_read(here::here(path, "TBEP-Bay_Segments.shp"),
                    quiet = TRUE)
wbids <- st_read(here::here(path, "Waterbody_IDs_(WBIDs).shp"),
                    quiet = TRUE)

# tbseg is a polygon layer included in tbeptools
# tidalcreeks is as well but needs to be subsetted; covers most of SWFL
```

```{r modify-shapefiles}
# pull out crs of the wbids for future joins, and make sure bay_segs is in that crs
to_crs <- st_crs(wbids)
bay_segs <- st_transform(bay_segs, crs = to_crs)

# subset wbids to only those that intersect bay segments
wbids <- st_intersection(wbids, bay_segs)
```

```{r}
yrsplit <- params$YeartoSplit
tslengthsplit <- params$nYearSegments

col_scale_divs <- tslengthsplit  # will actually be one category fewer than this, because of the sequence endpoints
```

This document will compare a hard download through the Water Quality Portal's website to the results from `tbeptools::read_importfib()`. The same figures and tables are produced for both download sources. They are *not* presented as separate tabs because that removed the functionality of a table of contents.   


# Website download  


### FIB Sampling points

This file contains *results* from microbiological sampling in the 5 TBEP counties from 1995-2022. We don't necessarily need to know what the results are, just that they exist. The query from Water Quality Portal was also set to only include stations with at least 10 sampling events. The latest download was performed 1/29/2024 and [this is the link to the query](https://www.waterqualitydata.us/#countrycode=US&statecode=US%3A12&countycode=US%3A12%3A057&countycode=US%3A12%3A081&countycode=US%3A12%3A101&countycode=US%3A12%3A103&countycode=US%3A12%3A105&sampleMedia=Water&characteristicType=Microbiological&startDateLo=01-01-1995&startDateHi=12-31-2022&minactivities=10&mimeType=csv&dataProfile=biological&providers=NWIS&providers=STEWARDS&providers=STORET).

That download was pre-processed using `R/WQP_preprocessing.R` to:  

-  filter out Groundwater samples - only keeping Surface Water.  
-  only keep sample types Fecal Coliform, Total Coliform, Enterococcus, and E. coli (eliminates DNA-based samples). Total Coliform is in the processed data frame but I am ignoring it in this exploration.  
-  only keep a small subset of columns relevant to this exploration.  

The pre-processing reduced the file size from 167MB to 40MB, and the smaller file is used in this document.  

```{r read-sample-info}
fib_samps <- readRDS(here::here("data-WQP", "biological_reduced.rds")) |> 
    filter(CharacteristicName %in% c("Fecal Coliform", "Enterococcus", "Escherichia coli"))
```

```{r, child = "_fib-expl-sourced.qmd"}

```

***
***

# `read_importfib` download  

just some test text for now








